\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2025-05-05 scratch/JSS-submission/Final_submission/jss5535.tex
create Final_submission/add files

\Abstract{
We describe the \pkg{cv} package, which implements cross-validation for
standard \proglang{R} regression models through a uniform and simple to
use generic function. Methods for the \code{cv()} function are provided
for many commonly employed classes of statistical models, including
mixed-effects models, and it is straightforward to extend the \pkg{cv}
package by writing methods for additional classes of models. The
\code{cv()} function also can cross-validate complex model-selection
procedures, such as those that include variable transformations,
predictor selection, and selection among competing models. The
\code{cv()} function generally supports parallel computations, and the
supplied methods for linear and generalized linear models take advantage
of computational efficiencies.
}

@@

\Address{
    John Fox\\
    McMaster University\\
    Hamilton, Ontario, Canada\\
  E-mail: \email{jfox@mcmaster.ca}\\
  URL: \url{https://www.john-fox.ca/}\\~\\
      Georges Monette\\
    York University\\
    Toronto, Ontario, Canada\\

@@

\usepackage{amsmath}

@@

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common: The data are randomly divided into two,
possibly equal-size, parts; the first part is used to develop and fit a
statistical model; and then the second part is used to assess the
adequacy of the model fit to the first part of the data. Data-division,
however, suffers from two problems: (1) Dividing the data decreases the
sample size and thus increases sampling error; and (2), even more
disconcertingly, the results can vary substantially based on the random
division of the data, particularly in smaller samples. See \citet[Sec.
5.3]{Harrell:2015} for this and other remarks about data-division and
cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are
randomly divided as equally as possible into several, say \(k\), parts,
called ``folds.'' The statistical model is fit \(k\) times, leaving each
fold out in turn. Each fitted model is then used to predict the response
variable for the cases in the omitted fold. A CV criterion (also termed
a ``cost'' or ``loss'' measure), such as the mean-squared error
(``MSE'') of prediction, is then computed using these predicted values.
In the extreme \(k = n\), the number of cases in the data, thus omitting
individual cases and refitting the model \(n\) times---a procedure
termed ``leave-one-out (LOO) cross-validation.''

Because the \(n\) models are each fit to \(n - 1\) cases, LOO CV
produces a nearly unbiased estimate of prediction error. The \(n\)
regression models are highly statistically dependent, however, based as
they are on nearly the same data, and so the resulting estimate of
prediction error has larger variance than if the predictions from the
models fit to the \(n\) data sets were independent.

Because predictions are based on smaller data sets, each of size
approximately \(n - n/k\), estimated prediction error for \(k\)-fold CV
with \(k = 5\) or \(10\) (commonly employed choices) is more biased than
estimated prediction error for LOO CV. It is possible, however, to
correct \(k\)-fold CV for bias (see Section
\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

The \pkg{cv} package for \proglang{R} automates the process of
cross-validation for standard \proglang{R} statistical model objects.
The principal function in the packages, also named \code{cv()}, has
methods for objects produced by a number of commonly employed
regression-modeling functions, including those for mixed-effects models:

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv", quietly=TRUE)
R> methods("cv")
\end{CodeInput}
\begin{CodeOutput}
[1] cv.default*  cv.function* cv.glm*      cv.glmmTMB*  cv.lm*      
[6] cv.lme*      cv.merMod*   cv.modList*  cv.rlm*     
see '?methods' for accessing help and source code
\end{CodeOutput}
\end{CodeChunk}

\begin{itemize}
\item
  The \code{"lm"} and \code{"glm"} methods are for linear and
  generalized-linear models fit respectively by the standard
  \proglang{R} \code{lm()} and \code{glm()} functions.
\item
  The \code{"modList"} method for \code{cv()} cross-validates several
  competing models, not necessarily of the same class, using the same
  division of the data into folds. The \code{cv()} function is
  introduced in the context of a preliminary example in Section
  \ref{preliminary-example-polynomial-regression} of the paper.
\item
  Cross-validating mixed-effects models, using the \code{"glmmTMB"},
  \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
  special considerations that we take up in Section
  \ref{cross-validating-mixed-effects-models}.
\item
  The \code{"function"} method for \code{cv()}, discussed in Section
  \ref{cross-validating-model-specification}, cross-validates a complex
  model-specification process that may, for example, involve choice of
  data transformations and predictors, or model selection via CV itself.
\item
  The \code{"default"} \code{cv()} method works (perhaps with a bit of
  coaxing) with many other existing regression-model classes for which
  there is an \code{update()} method that accepts a \code{data}
  argument. More generally, the \pkg{cv} package is designed to be
  extensible, as discussed in Section \ref{extending-the-cv-package}.
\end{itemize}

A number of existing \proglang{R} packages include functions for
cross-validating regression models. We briefly situate the \pkg{cv}
package relative to other \proglang{R} software for cross-validation in
Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.

The final section of the paper (Section \ref{computational-notes})
describes some computational details, including efficient CV
computations for linear and generalized-linear models and computation of
bias-corrected CV criteria.

In the interest of brevity, we won't describe all of the features of the
\pkg{cv} package here, concentrating on the aspects of the package that
are relatively novel. For example, the \code{cv()} function can perform
computations in parallel and can independently replicate a
cross-validation procedure several times. There are also data management
facilities in the package, such as coercing the objects produced by the
\code{cv()} function into data frames for further analysis. These and
other features not discussed in this paper are taken up in the vignettes
distributed with the package, which also provides greater detail on some
of topics that we do describe, such as extensions to the package.

\section{Preliminary Example: Polynomial
regression}\label{preliminary-example-polynomial-regression}

The data for the example in this section are drawn from the \pkg{ISLR2}
package for \proglang{R}, associated with \citet{JamesEtAl:2021}. The
presentation here is close (though not identical) to that in the
original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
demonstrates the use of the \texttt{cv()} function.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \pkg{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function. The \pkg{boot} package is part of the
  standard \proglang{R} distribution.}

The \code{Auto} dataset contains information about 392 cars:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Auto", package="ISLR2")
R> summary(Auto)
\end{CodeInput}
\begin{CodeOutput}
      mpg          cylinders      displacement     horsepower        weight    
 Min.   : 9.00   Min.   :3.000   Min.   : 68.0   Min.   : 46.0   Min.   :1613  
 1st Qu.:17.00   1st Qu.:4.000   1st Qu.:105.0   1st Qu.: 75.0   1st Qu.:2225  
 Median :22.75   Median :4.000   Median :151.0   Median : 93.5   Median :2804  
 Mean   :23.45   Mean   :5.472   Mean   :194.4   Mean   :104.5   Mean   :2978  
 3rd Qu.:29.00   3rd Qu.:8.000   3rd Qu.:275.8   3rd Qu.:126.0   3rd Qu.:3615  
 Max.   :46.60   Max.   :8.000   Max.   :455.0   Max.   :230.0   Max.   :5140

  acceleration        year           origin                      name    
 Min.   : 8.00   Min.   :70.00   Min.   :1.000   amc matador       :  5  
 1st Qu.:13.78   1st Qu.:73.00   1st Qu.:1.000   ford pinto        :  5  
 Median :15.50   Median :76.00   Median :1.000   toyota corolla    :  5  
 Mean   :15.54   Mean   :75.98   Mean   :1.577   amc gremlin       :  4  
 3rd Qu.:17.02   3rd Qu.:79.00   3rd Qu.:2.000   amc hornet        :  4  
 Max.   :24.80   Max.   :82.00   Max.   :3.000   chevrolet chevette:  4  
                                                 (Other)           :365  
\end{CodeOutput}
\end{CodeChunk}

With the exception of \texttt{origin} (which we don't use here), these
variables are largely self-explanatory, except possibly for units of
measurement: for details see \texttt{help("Auto",\ package="ISLR2")}.

@@

{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-polynomials-1} }\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-polynomials-2} }

@@

\caption{(a) \code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5. (b) Estimated squared error as a function of polynomial degree, $p$.}\label{fig:mpg-horsepower-polynomials}
\end{figure}
\end{CodeChunk}

We'll focus here on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example the scatterplot for log-transformed
  \texttt{mpg} and \texttt{horsepower}, produced by
  \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
  (execution of which is left to the reader). We revisit the \code{Auto}
  data in Section \ref{cross-validating-model-specification}.}
Polynomial fits for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
clearly inappropriate; the fits for \(p = 2\) (quadratic) through \(4\)
are very similar; and the fit for \(p = 5\) may over-fit the data by
chasing one or two relatively high \texttt{mpg} values at the right (but
see the CV results reported below).

Figure \ref{fig:mpg-horsepower-MSE-se} shows two measures of estimated
(squared) error as a function of polynomial-regression degree: The
mean-squared error (``MSE''), defined as
\(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
the usual residual variance, defined as
\(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
The former necessarily declines with \(p\) (or, more strictly, can't
increase with \(p\)), while the latter gets slightly larger for the
largest values of \(p\), with the ``best'' value, by a small margin, for
\(p = 7\).

The generic \texttt{cv()} function has an \texttt{"lm"} method,

@@

\begin{itemize}
\item
  \code{model}, an \code{"lm"} object, the only required argument.
\item
  \code{data}, which can usually be inferred from the \code{model}
  object.
\item
  \code{criterion}, a function to compute the CV criterion (defaulting
  to \code{mse}).
\item
  \code{k}, the number of folds to employ (defaulting to \code{10}); the
  character value \code{"n"} or \code{"loo"} may be supplied to specify
  leave-one-out cross-validation.
\item
  \code{reps}, the number of times to repeat the CV procedure
  (defaulting to 1).
\item
  \code{seed}, the seed for \proglang{R}'s pseudo-random number
  generator; if not specified a value is randomly selected, reported,
  and saved, so that the CV procedure is replicable.
\item
  \code{confint}, whether or not to compute a confidence interval for
  the CV criterion, defaulting to \code{TRUE} if there are at least 400
  cases; a confidence interval is computed only if the CV criterion can
  be expressed as the average of casewise components (see Section
  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
  for details).
\item
  \code{level}, the level for the confidence interval (defaulting to
  \code{0.95}).
\item
  \code{method}, the computational method to employ: \code{"hatvalues"}
  is relevant only for LOO CV and bases computation on the hatvalues for
  the linear model; \code{"Woodbury"} employs the Woodbury matrix
  identity to compute the CV criterion with each fold deleted;
  \code{"naive"} updates the model using the \code{update()} function;
  and \code{"auto"} selects \code{"hatvalues"} for LOO CV and
  \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
  efficient than literally updating the least-squares fit (see below and
  Section
  \ref{efficient-computations-for-linear-and-generalized-linear-models}).
\item
  \code{ncores}, the number of cores to employ for parallel computation;
  if \code{cores = 1} (the default), the computations are not
  parallelized.
\end{itemize}

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to
the \code{Auto} data:

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
R> summary(m.auto)
\end{CodeInput}
\begin{CodeOutput}

Call:
lm(formula = mpg ~ poly(horsepower, 2), data = Auto)

Residuals:
     Min       1Q   Median       3Q      Max 
-14.7135  -2.5943  -0.0859   2.2868  15.8961

Coefficients:
                      Estimate Std. Error t value Pr(>|t|)
(Intercept)            23.4459     0.2209  106.13   <2e-16
poly(horsepower, 2)1 -120.1377     4.3739  -27.47   <2e-16
poly(horsepower, 2)2   44.0895     4.3739   10.08   <2e-16

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 180983
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.37365
bias-adjusted cross-validation criterion = 19.35319
95% CI for bias-adjusted CV criterion = (15.87805, 22.82834)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

The function reports the CV estimate of MSE, a bias-adjusted estimate of
the MSE (the bias adjustment is explained in Section
\ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}),
and the MSE is also computed for the original, full-sample regression.
Because the number of cases \(n = 392 < 400\) for the \code{Auto} data,
we set the argument \code{confint=TRUE} to obtain a confidence interval
for the MSE, which proves to be quite wide.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo")
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: hatvalues
criterion: mse
cross-validation criterion = 19.24821
\end{CodeOutput}
\end{CodeChunk}

The \code{"hatvalues"} method reports only the CV estimate of MSE.
Alternative methods are to use the Woodbury matrix identity or the
``naive'' approach of literally refitting the model with each case
omitted. All three methods produce exact results for a linear model
(within the precision of floating-point computations):

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="naive", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: naive
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

This is a small regression problem and all three computational
approaches are essentially instantaneous, but it is still of interest to
investigate their relative speed. In the following comparison, we
include the \code{cv.glm()} function from the \pkg{boot} package, which
takes the naive approach, and for which we have to fit the linear model
as an equivalent Gaussian GLM. We use the \code{microbenchmark()}
function from the package of the same name \citep{Mersmann:2023} for the
timings:

>>>>

\Abstract{
We describe the \pkg{cv} package, which implements cross-validation for
standard \proglang{R} regression models through a uniform and simple to
use generic function. Methods for the \code{cv()} function are provided
for many commonly employed classes of statistical models, including
mixed-effects models, and it is straightforward to extend the \pkg{cv}
package by writing methods for additional classes of models. The
\code{cv()} function can also cross-validate complex model-selection
procedures, such as those that include variable transformations,
predictor selection, and selection among competing models. The
\code{cv()} function generally supports parallel computations, and the
supplied methods for linear and generalized linear models take advantage
of computational efficiencies.
}

@@

\Address{
    John Fox\\
    McMaster University\\
    Hamilton, Ontario, Canada\\
  E-mail: \email{jfox@mcmaster.ca}\\
  URL: \url{https://www.john-fox.ca/}\\~\\
      Georges Monette\\
    York University\\
    Toronto, Ontario, Canada\\
  E-mail: \email\\
  URL: \url{http://blackwell.math.yorku.ca/gmonette}\\~\\
  }

@@

\usepackage{amsmath} \usepackage{amssymb}

@@

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common \citep[see, e.g.,][]{Barnard:1974}: The
data are randomly divided into two, possibly equal-size, parts; the
first part is used to develop and fit (in the language of machine
leaning, ``train'') a statistical model; and then the second part is
used to assess the adequacy of (``validate'') the model fit to the first
part of the data. Data-division, however, suffers from two problems: (1)
Dividing the data decreases the sample size and thus increases sampling
error; and (2), even more disconcertingly, the results can vary
substantially based on the random division of the data, particularly in
smaller samples. See \citet[Sec. 5.3]{Harrell:2015} for cogent remarks
about data-division and cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are
randomly divided as equally as possible into several, say \(k\), parts,
called ``folds.'' The statistical model is fit \(k\) times, leaving each
fold out in turn. Each fitted model is then used to predict the response
variable for the cases in the omitted fold. A CV criterion (also termed
a ``cost'' or ``loss'' measure), such as the mean-squared error
(``MSE'') of prediction, is then computed using these predicted values.
In the extreme, \(k\) equals \(n\), the number of cases in the data,
thus omitting individual cases and refitting the model \(n\) times---a
procedure termed ``leave-one-out (LOO) cross-validation.''

Because the \(n\) models are each fit to \(n - 1\) cases, LOO CV
produces a nearly unbiased estimate of the prediction error compared
with a model fit to \(n\) cases. The \(n\) regression models are highly
statistically dependent, however, based as they are on nearly the same
data, and so the resulting estimate of prediction error has larger
variance than if the predictions from the models fit to the \(n\) data
sets were independent.

Because predictions are based on smaller data sets, each of size
approximately \(n - n/k\), estimated prediction error for \(k\)-fold CV
with \(k = 5\) or \(10\) (commonly employed choices) is more biased than
estimated prediction error for LOO CV. It is possible, however, to
correct \(k\)-fold CV for bias (see Appendix A.2).

@@

The \pkg{cv} package for \proglang{R} automates the process of
cross-validation for standard \proglang{R} statistical models. The
principal function in the package, also named \code{cv()}, has methods
for objects produced by a number of commonly employed
regression-modeling functions, including those for mixed-effects models:

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv", quietly = TRUE)
R> methods("cv")
\end{CodeInput}
\begin{CodeOutput}
[1] cv.default*  cv.function* cv.glm*      cv.glmmTMB*  cv.lm*
[6] cv.lme*      cv.merMod*   cv.modList*  cv.rlm*
see '?methods' for accessing help and source code
\end{CodeOutput}
\end{CodeChunk}

The \code{cv()} function is introduced in the context of a preliminary
example in Section \ref{preliminary-example-polynomial-regression} of
the paper.

\begin{itemize}
\item
  The \code{"lm"} and \code{"glm"} methods are for linear and
  generalized-linear models fit respectively by the standard
  \proglang{R} \code{lm()} and \code{glm()} functions.
\item
  The \code{"modList"} method for \code{cv()} cross-validates several
  competing models, not necessarily of the same class, using the same
  division of the data into folds.
\item
  Cross-validating mixed-effects models, using the \code{"glmmTMB"},
  \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves
  special considerations that we take up in Section
  \ref{cross-validating-mixed-effects-models}.
\item
  The \code{"function"} method for \code{cv()}, discussed in Section
  \ref{cross-validating-model-specification}, cross-validates a
  potentially complex model-specification process that may, for example,
  involve choice of data transformations and predictors, or model
  selection via CV itself.
\item
  The \code{"default"} \code{cv()} method works (perhaps with a bit of
  coaxing) with many other existing regression-model classes for which
  there is an \code{update()} method that accepts a \code{data}
  argument. More generally, the \pkg{cv} package is designed to be
  extensible, as discussed in Section \ref{extending-the-cv-package}.
\end{itemize}

A number of existing \proglang{R} packages include functions for
cross-validating regression models. In Section
\ref{comparing-cv-to-other-software-for-cross-validation}, we briefly
situate the \pkg{cv} package relative to other \proglang{R} software for
cross-validation (directly comparing \pkg{cv} to the \pkg{rsample} and
\pkg{caret} packages), and to other statistical software (in particular,
\proglang{SAS}, \proglang{Stata}, and \proglang{Python}).

An appendix to the paper describes some computational details, including
efficient CV computations for linear and generalized-linear models, and
computation of bias-corrected CV criteria.

In the interest of brevity, we won't describe all of the features of the
\pkg{cv} package here, concentrating on the aspects of the package that
are relatively novel. For example, the \code{cv()} function can perform
computations in parallel and can independently replicate a
cross-validation procedure several times. There are also data-management
facilities in the package, such as coercing the objects produced by the
\code{cv()} function into data frames for further analysis. These and
other features not discussed in this paper are taken up in the vignettes
distributed with the package, which also provide greater detail on some
topics that we do describe, such as extensions to the package.

\section{Preliminary example: Polynomial
regression}\label{preliminary-example-polynomial-regression}

The data for the example in this section are drawn from the \pkg{ISLR2}
package for \proglang{R}, associated with \citet{JamesEtAl:2021}. The
presentation here is close (though not identical) to that in the
original source \citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it
demonstrates the use of the \code{cv()} function.\footnote{\citet{JamesEtAl:2021}
  use the \code{cv.glm()} function in the \pkg{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \code{cv.glm()} is an independent function and not a method of a
  \code{cv()} generic function. The \pkg{boot} package is part of the
  standard \proglang{R} distribution.}

The \code{Auto} data set contains information about 392 cars:

\begin{CodeChunk}
\begin{CodeInput}
R> data("Auto", package = "ISLR2")
R> summary(Auto)
\end{CodeInput}
\begin{CodeOutput}
      mpg         cylinders     displacement   horsepower        weight
 Min.   : 9.0   Min.   :3.00   Min.   : 68   Min.   : 46.0   Min.   :1613
 1st Qu.:17.0   1st Qu.:4.00   1st Qu.:105   1st Qu.: 75.0   1st Qu.:2225
 Median :22.8   Median :4.00   Median :151   Median : 93.5   Median :2804
 Mean   :23.4   Mean   :5.47   Mean   :194   Mean   :104.5   Mean   :2978
 3rd Qu.:29.0   3rd Qu.:8.00   3rd Qu.:276   3rd Qu.:126.0   3rd Qu.:3615
 Max.   :46.6   Max.   :8.00   Max.   :455   Max.   :230.0   Max.   :5140

  acceleration       year        origin                     name
 Min.   : 8.0   Min.   :70   Min.   :1.00   amc matador       :  5
 1st Qu.:13.8   1st Qu.:73   1st Qu.:1.00   ford pinto        :  5
 Median :15.5   Median :76   Median :1.00   toyota corolla    :  5
 Mean   :15.5   Mean   :76   Mean   :1.58   amc gremlin       :  4
 3rd Qu.:17.0   3rd Qu.:79   3rd Qu.:2.00   amc hornet        :  4
 Max.   :24.8   Max.   :82   Max.   :3.00   chevrolet chevette:  4
                                            (Other)           :365
\end{CodeOutput}
\end{CodeChunk}

With the exception of \code{origin} (which we don't use here), these
variables are largely self-explanatory, except possibly for units of
measurement: For details see \code{help("Auto",}
\code{package = "ISLR2")}.

@@

{\centering \subfloat[\label{fig:mpg-horsepower-polynomials-1}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-1} }~~~\subfloat[\label{fig:mpg-horsepower-polynomials-2}]{\includegraphics[width=0.45\linewidth]{Figures/mpg-horsepower-polynomials-2} }

@@

\caption{(a) \code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5. (b) Estimated squared error as a function of polynomial degree, $p = 1, \ldots, 10$.}\label{fig:mpg-horsepower-polynomials}
\end{figure}
\end{CodeChunk}

We'll focus on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in
Figure~\ref{fig:mpg-horsepower-polynomials} (a). The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example, the scatterplot for
  log-transformed \code{mpg} and \code{horsepower}, produced by
  \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of
  which is left to the reader). We revisit the \code{Auto} data in
  Section \ref{cross-validating-model-specification}.} Polynomial fits
for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly
inappropriate; the fits for \(p = 2\) (quadratic) through \(4\) are very
similar; and the degree-5 polynomial may over-fit the data by chasing
one or two relatively high \code{mpg} values at the right (but see the
CV results reported below).

Figure \ref{fig:mpg-horsepower-polynomials} (b) shows two measures of
estimated (squared) error as a function of polynomial-regression degree:
The mean-squared error (``MSE''), defined as
\(\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2\), and
the usual residual variance, defined as
\(\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2\).
The former necessarily declines with \(p\) (or, more strictly, can't
increase with \(p\)), while the latter gets slightly larger for the
largest values of \(p\), with the ``best'' value, by a small margin, for
\(p = 7\).

The generic \code{cv()} function has an \code{"lm"} method,

@@

\begin{itemize}
\item
  \code{model}, an \code{"lm"} object, the only required argument.
\item
  \code{data}, the data to which the model was fit, and which can
  usually be inferred from the \code{model} object, using the
  \code{get_data()} function from the \pkg{insight} package
  \citep{LudeckeWaggonerMakowski:2019}. The \pkg{insight} package
  provides a robust way to access model components for many classes of
  statistical models.
\item
  \code{criterion}, a function to compute the CV criterion (defaulting
  to \code{mse}).
\item
  \code{k}, the number of folds to employ (defaulting to \code{10}); the
  character value \code{"n"} or \code{"loo"} may be supplied to specify
  leave-one-out cross-validation.
\item
  \code{reps}, the number of times to repeat the CV procedure
  (defaulting to 1).
\item
  \code{seed}, the seed for \proglang{R}'s pseudo-random number
  generator; if not specified a value is randomly selected, reported,
  and saved, so that the CV procedure is replicable.
\item
  \code{confint}, whether or not to compute a confidence interval for
  the CV criterion, defaulting to \code{TRUE} if there are at least 400
  cases; a confidence interval is computed only if the CV criterion can
  be expressed as the average of casewise components (see Appendix A.2
  for details).
\item
  \code{level}, the level for the confidence interval (defaulting to
  \code{0.95}).
\item
  \code{method}, the computational method to employ: \code{"hatvalues"}
  is relevant only for LOO CV and bases computation on the hatvalues for
  the linear model; \code{"Woodbury"} employs the Woodbury matrix
  identity to compute the CV criterion with each fold deleted;
  \code{"naive"} updates the model using the \code{update()} function;
  and \code{"auto"} (the default) selects \code{"hatvalues"} for LOO CV
  and \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
  efficient than directly recomputing the least-squares fit (see below
  and Appendix A.1).
\item
  \code{ncores}, the number of cores to employ for parallel computation;
  if \code{ncores = 1} (the default), the computations are not
  parallelized.
\end{itemize}

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to
the \code{Auto} data:\footnote{We load the \pkg{car} package
  \citep{FoxWeisberg:2019} here for the \code{brief()} function and to
  use below.}

\begin{CodeChunk}
\begin{CodeInput}
R> library("car", quietly = TRUE)
R>
R> m.auto <- lm(mpg ~ poly(horsepower, 2), data = Auto)
R> brief(m.auto)
\end{CodeInput}
\begin{CodeOutput}
           (Intercept) poly(horsepower, 2)1 poly(horsepower, 2)2
Estimate        23.446              -120.14                44.09
Std. Error       0.221                 4.37                 4.37

 Residual SD = 4.37 on 389 df, R-squared = 0.688
\end{CodeOutput}
\begin{CodeInput}
R> (cv.auto <- cv(m.auto, confint = TRUE))
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 623225
\end{CodeOutput}
\begin{CodeOutput}
cross-validation criterion (mse) = 19.409
\end{CodeOutput}
\begin{CodeInput}
R> summary(cv.auto)
\end{CodeInput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.409
bias-adjusted cross-validation criterion = 19.386
95% CI for bias-adjusted CV criterion = (15.885, 22.887)
full-sample criterion = 18.985
\end{CodeOutput}
\end{CodeChunk}

The \texttt{print()} method for \texttt{"cv"} objects reports the CV
estimate of MSE; the \texttt{summary()} method, which we'll use
extensively in the sequel, in addition reports a bias-adjusted estimate
of the MSE (the bias adjustment is explained in Appendix A.2), and the
MSE is also computed for the original, full-sample regression. Because
the number of cases \(n = 392 < 400\) for the \code{Auto} data, we set
the argument \code{confint = TRUE} to obtain a confidence interval for
the MSE, which proves to be quite wide.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(cv(m.auto, k = "loo"))
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: hatvalues
criterion: mse
cross-validation criterion = 19.248
\end{CodeOutput}
\end{CodeChunk}

The \code{"hatvalues"} method computes only the CV estimate of MSE.
Alternative methods are to use the Woodbury matrix identity or the
``naive'' approach of literally refitting the model with each case
omitted. All three methods produce exact results for a linear model
(within the precision of floating-point computations); for example:

\begin{CodeChunk}
\begin{CodeInput}
R> summary(cv(m.auto, k = "loo", method = "naive", confint = TRUE))
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: naive
criterion: mse
cross-validation criterion = 19.248
bias-adjusted cross-validation criterion = 19.248
95% CI for bias-adjusted CV criterion = (15.779, 22.717)
full-sample criterion = 18.985
\end{CodeOutput}
\end{CodeChunk}

@@

This is a small regression problem and all three computational
approaches are essentially instantaneous, but it is still of interest to
investigate their relative speed. In the following comparison, we
include the \code{cv.glm()} function from the \pkg{boot} package, which
takes the naive approach, and for which we have to fit the linear model
as an equivalent Gaussian GLM; the output shown is for the CV estimate
of MSE and the bias-corrected CV estimate. We use the
\code{microbenchmark()} function from the package of the same name
\citep{Mersmann:2023} for the timings:

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data = Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.248 19.248
\end{CodeOutput}
\begin{CodeInput}
R> set.seed(19412)
R> print(microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k = "loo"),
+   Woodbury = cv(m.auto, k = "loo", method = "Woodbury"),
+   naive = cv(m.auto, k = "loo", method = "naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times = 10, unit = "relative"), signif = 3)
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: relative
      expr    min     lq   mean median     uq   max neval cld
 hatvalues   1.00   1.00   1.00   1.00   1.00  1.00    10 a
  Woodbury   7.47   7.15   5.77   6.74   6.62  2.35    10 a
     naive 124.00 119.00  98.20 112.00 111.00 47.50    10  b
    cv.glm 217.00 207.00 172.00 197.00 198.00 75.20    10   c
\end{CodeOutput}
\end{CodeChunk}

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.\footnote{Out of
  impatience, we asked \code{microbenchmark()} to execute each command
  only 10 times rather than the default 100. With the exception of the
  last column, the output is self-explanatory. The last column shows
  which methods have average timings that are statistically
  distinguishable (as indicated by different letters). Because of the
  small number of repetitions (i.e., 10), the \code{"hatvalues"} and
  \code{"Woodbury"} methods aren't distinguishable, but the difference
  between these methods persists when we perform more repetitions---we
  invite the patient reader to redo this computation with the default
  \code{times = 100} repetitions.}
